---
title: Fourth Graph - How has the MOMA gotten its paintings over the years?
author: Matt Pomaville
date: '2018-06-13'
slug: fourth-graph
categories:
  - CS631
  - R
tags:
  - ggplot
description: Desc
hacker_news_id: ''
lobsters_id: ''
meta_img: /images/image.jpg
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This graph looks at the number of gifted paintings vs the number of paintings acqired through purchase or exchange by the MOMA grouped by decade.

##Type of Graph
This graph is a stacked column graph.

## The Data
This data came from a pre-cleaned version of the Museum of Museum Art data set generated by www.fivethirtyeight.com. It was cleaned by Alison Hill and further modified by Matt Pomaville. It is available [here]("https://raw.githubusercontent.com/Dystroglycan1234/Data-Vis-Class/master/artworks-cleaned.csv").

This dataset has a lot of information but I am particularly interested in two columns:

* year_acquired - The year the MOMA got the painting
* gift - A binary value (TRUE or FALSE) indicating if the painting was a gift or a prchase/exchange.

##Audience
This graph is meant for anyone interested in the art at the MOMA in New York. It is easy to interpret and doesn't require much background knowledge to understand.

## What is the graph showing?
This graph shows the number of paintings acquired by the MOMA grouped by decade. The decade columns are split into gift and non-gift paintings. There doesn't seem to be any trend regarding how paintings are acquired.

##How to read it
Reading a stacked column graph is similar to reading any other column graph. Look for any trend in the y-values as the x-value increases. There are multiple meaningful trends across the x-axis, including noise or no trend.

##Presentation and Annotation
There is little annotation in this graph. If there was a particular reason a decade was important above the others an annotation coule be added to denote this.

##Variations
There are not many alternative graphs that coule be used to show this data in the same format. A boxplot of each decade using the mean of each year as a datapoint could be used to make the same point.

##Methods
Call the required libraries: `tidyverse` and `wesanderson`.

```{r message = FALSE}
library(tidyverse)
library(wesanderson)
```

Pull the data from github.

```{r}
moma <- read_csv("https://raw.githubusercontent.com/Dystroglycan1234/Data-Vis-Class/master/artworks-cleaned.csv")
  
```

Summarise the data into three columns:

* gift - A binary factor showing if the painting was a gift or not.
* decade_acquired - the decade the painting was acquired in.
* total - The total number of paintings acquired by a given means in a year.

```{r}
moma_gift <- moma %>% 
  group_by(gift) %>% 
  count(year_acquired) %>% 
  mutate(decade_acquired = signif(year_acquired, digits = 3)) %>% 
  ungroup() %>% 
  group_by(gift, decade_acquired) %>% 
  summarize(sum(n))

names(moma_gift) <- c("gift", "decade_acquired", "total")
```

Graph the data as a column graph with year on the x-axis, number acquired on the y-axis, and the stacked columns colored by whether or not the painting was a gift.

```{r warning = FALSE}

gift_plot_bydecade <- ggplot(moma_gift, aes(x = decade_acquired, y = total)) +
  geom_col(aes(fill = gift)) +
  ggtitle("Have the number of gifts to MoMA increased")+
  labs(x = "Decade Acquired", y = "Number of paintings acquired") +
  scale_fill_manual(name = "Gift", values = c("#0B775E", "#F2300F"), labels = c("Purchase or Exchange", "Gift"))
  

gift_plot_bydecade
```
